The hydra and the rooster

### 2020-08-12 latex/chapter-hydras.tex
proof simplification in L_alpha

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}
\index{Coq!Techniques!Type classes}

@@

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#Hvariant}{Hydra.Hydra\_Definitions}}
\index{Coq!Techniques!Type classes}
\label{sect:hvariant-def}

@@

>>>>

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}
\index{Coq!Type classes}

@@

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#Hvariant}{Hydra.Hydra\_Definitions}}

\index{Coq!Type classes}
\label{sect:hvariant-def}

@@

\label{remark-infix}
\begin{remark}
We use in \coq{} scripts the infix notations \verb@o<@ and \verb@o<=@ for
the orders $<$ and $\leq$ on ordinals. This convention allows us to define
implicit coercions from \texttt{nat} into various types for representing ordinals.
For instance a proposition like \verb@ 6 o< 8@ will be interpreted as 
``the finite ordinal 6 is strictly less than the finite ordinal 8'', whereas 
\verb@ 6 o< 8@ is a statement about natural numbers. Since many lemmas talk about both ordinals and natural numbers, we wished to avoid too many scope changes inside propositions.
\end{remark}

### 2020-08-12 latex/chapter-large-sets.tex
proof simplification in L_alpha

\section{Introduction}

@@

We express this number in terms of the Hardy hierarchy of fast-growing 
functions~\cite{BW85, Wainer1970, KS81, Promel2013}.
 From the \coq{} user's point of view, such  functions are  very 
attractive:  they are defined as functions  in \gallina{}, and we can apply rhem \emph{in theory}, but they are so complex that you will never be able to look at the result of the computation.
 Thus, our knowledge on these functions must rely on \emph{proofs}. In our development, we use often the rewriting rules generated by \coq's \texttt{Equations} plug-in.

@@

\begin{definition}
Let $0<\alpha<\epsilon_0$ be any ordinal, and $s$ a sequence of strictly positive natural numbers. We say that $s$ is an  \emph{$\alpha$-large} 
sequence  if $s$ determines a path to the ordinal $0$.
We say that $s$ is \emph{minimally $\alpha$-large} (abbreviated in
\emph{$\alpha$-mlarge} if $s$ if $s$ is $\alpha$-large 
 and every strict prefix of $s$ leads to a non-zero ordinal.

@@

\begin{remark}
  Ketonen and Solovay~\cite{KS81} consider  large finite \emph{sets} of natural numbers,  but they are mainly used as sequences. Thus, we chosed to represent them explicitely as (sorted) lists. 
\end{remark}

@@

\index{Predicates!mlarge @ mlarge (minimal large sequences)}
\begin{Coqsrc}
Definition mlarge alpha s := path_to zero s alpha.
\end{Coqsrc}

@@

This verification can be done by a simple computation. First, we define a function for following a path, by successively computing canonical sequences.

\begin{Coqsrc}
Fixpoint gnaw (alpha : T1) (s: list nat) :=
  match s with
    | nil => alpha
    | (0::s') => gnaw  alpha s'
    | (S i :: s')  =>  gnaw (canonS i alpha) s'
  end.
\end{Coqsrc}

@@

For instance, we can verify that the interval $[6,509]$ is $\omega^2$-mlarge.

@@

@@

First, we notice that this  number $l$ exists, thanks to the following theorems:

\begin{Coqsrc}
canonS_LT:
  forall (i : nat) (alpha : T1),
  nf alpha -> alpha <> zero -> canonS alpha i o< alpha.

T1_wf: well_founded LT.
\end{Coqsrc}

Moreover, it is unique:

\begin{Coqsrc}
Lemma mlarge_unicity alpha k l l' : 
  mlarge alpha (interval (S k) l) ->
  mlarge alpha (interval (S k) l') ->
  l = l'.
\end{Coqsrc}

Thus, it seems obvious that there should be a function, parameterized by $\alpha$ which associates to any  strictly positive integer $k$ the number $l$ such that
the interval $[k,l]$ is $\alpha$-large. It would be fine to write in \gallina{} a definition like this:

@@

But we do not know how to fill the dots yet \dots{}   In the next section, we will 
use \coq{} to reason  about the \emph{specification} of \texttt{L},
prove properties of any function which satisfies this specification, and finally define a correct function with the help of the \texttt{coq-equations} plug-in.

\section{Looking for a function to define}

Let $0<\alpha<\epsilon_0$ be an ordinal term. We consider the functions which associate to each stritly positive integer $k$ the number $l$, where 
the interval $[k,l-1]$\;\footnote{also denoted by $[k,l)$}  is $\alpha$-mlarge.

@@

Let us now prove properties of any function $f$ which satisfies 
\texttt{L\_spec}. We are looking for properties which could be used for writing \emph{equations} and prove the correctness of the function generated by the \texttt{coq-equations} plugin. Moreover, they will give us some examples of
$L_\alpha$ for small values of $\alpha$.

Our exploration of the $L_\alpha$s will follow the usual scheme : transfinite induction, and proof-by-cases : zero, successors and limit ordinals.

@@

Let $\beta$ be some ordinal, and assume the arithmetic function $f$ satisfies 
the specification $\texttt{L\_spec}\;\beta$.  Let $k$ be any natural number.
Any path from $\texttt{succ}\,\beta$ to $0$ starting at $k+1$ can be decomposed into a first step from $\texttt{succ}\,\beta$ to $\beta$, then a path from
$\beta$ at $k+2$ to $0$. 
By hypothesis the interval $[k+2, f(k+2)-1]$ is $\beta$-mlarge.

If we assume the inequality $k+1\leq f(k+2)$, we infer that the interval 
$[k+1, f(k+2)]$ is $\beta+1$-mlarge. This new hypothesis allows us 
to apply the following lemma 
(from~\href{../src/html/hydras.Prelude.MoreLists.html\#interval_app}{Prelude.MoreLists}).

\begin{Coqsrc}
Lemma interval_app (i j k:nat):
  (i <= j)%nat -> (j <= k)%nat ->
  interval i k = interval i j ++ interval (S j) k.
\end{Coqsrc}

For ensuring this property, we assume that $f\;k$ is always greater than $k$, which we note \texttt{S <= f}, or \texttt{(fun\_le S f)} (defined 
in~\href{../src/html/hydras.Prelude.Iterates.html\#fun_le}{Prelude.Iterates}).

@@

\section{Using \texttt{Equations}}
\label{sect:equations}

@@

@@

\section{The Wainer-Hardy hierarchy}

In order to give an idea of the complexity of the functions  $L_\alpha$s, we compare them with a better known family of functions, the so called \emph{Wainer-Hardy hierarchy} of fast growing functions,
presented for instance in~\cite{Promel2013}. 
\index{Maths!Hardy Hierarchy}

@@

\section{The Wainer hierarchy (functions \texorpdfstring{$F_\alpha$}{F\_alpha})}

>>>>

%\section{Introduction}

@@

We express this number in terms of the Hardy hierarchy of fast-growing 
functions~\cite{BW85, Wainer1970, KS81, Promel2013}.
 From the \coq{} user's point of view, such  functions are  very 
attractive:  they are defined as functions  in \gallina{}, and we can apply them \emph{in theory}, but they are so complex that you will never be able to look at the result of the computation.
 Thus, our knowledge on these functions must rely on \emph{proofs}. In our development, we use often the rewriting rules generated by \coq's \texttt{Equations} plug-in.

@@

\subsection{Definition}

\begin{definition}
Let $0<\alpha<\epsilon_0$ be any ordinal, and $s$ a sequence of strictly positive natural numbers. 
We say that $s$ is \emph{minimally $\alpha$-large} (in short:
\emph{$\alpha$-mlarge}) if $s$ if $s$ is $\alpha$-large 
 and every strict prefix of $s$ leads to a non-zero ordinal (\emph{cf} Sect.~\vref{sect:path-to-def}).

@@

@@

\index{Predicates!mlarge @ mlarge (minimal large sequences)}
\begin{Coqsrc}
Definition mlarge alpha (s:list nat) := path_to zero s alpha.
\end{Coqsrc}

\begin{remark}
  Ketonen and Solovay~\cite{KS81} consider  large finite \emph{sets} of natural numbers,  but they are mainly used as sequences. Thus, we chosed to represent them explicitely as (sorted) lists.

They also consider large (but not minimally large) sets. They can be defined in
\coq{} as follows:

\begin{Coqsrc}
Fixpoint gnaw (alpha : T1) (s: list nat) :=
  match s with
    | nil => alpha
    | (0::s') => gnaw  alpha s'
    | (S i :: s')  =>  gnaw (canonS i alpha) s'
  end.

Definition large alpha (s:list nat) := gnaw alpha s = zero.
\end{Coqsrc}
\end{remark}

@@

\subsection{Example}

@@

For instance, we can verify that the interval $[6,510]$ is $\omega^2$-mlarge.

@@

\section{Computing mlarge sequences}

@@

First, we notice that this  number $l$ exists, since the segment $[0,\epsilon_0)$ is well-founded and $\canonseq{\alpha}{i}<\alpha$ for any $i$ and $\alpha>0$.
Moreover, it is unique:

\vspace{4pt}
\noindent
\emph{From Module~\href{../src/html/hydras.Epsilon0.Large_Sets.html}{Epsilon0.Large\_Sets}}
\begin{Coqsrc}
Lemma mlarge_unicity alpha k l l' : 
  mlarge alpha (interval (S k) l) ->
  mlarge alpha (interval (S k) l') ->
  l = l'.
\end{Coqsrc}

Thus, it seems obvious that there must exist a function, parameterized by $\alpha$ which associates to any  strictly positive integer $k$ the number $l$ such that
the interval $[k,l]$ is $\alpha$-mlarge. It would be fine to write in \gallina{} a definition like this:

@@

But we do not know how to fill the dots yet \dots{}   In the next section, we will 
use \coq{} to reason  about the \emph{specification} of \texttt{L},
prove properties of any function which satisfies this specification.
In Sect.~\ref{sect:L-equaions}, we use the \texttt{coq-equations} plug-in
to define a function \texttt{L\_}, and prove its correctness w.r.t. its specification.

\subsection{Looking for a function definition}

Let $0<\alpha<\epsilon_0$ be an ordinal term. We consider the functions which associate to each stritly positive integer $k$ the number $l$, where 
the interval $[k,l)$ is $\alpha$-mlarge.

@@

Let us now prove properties of any function $f$ (if any) which satisfies 
\texttt{L\_spec}. We are looking for properties which could be used for writing \emph{equations} and prove the correctness of the function generated by the \texttt{coq-equations} plugin. Moreover, they will give us some examples of
$L_\alpha$ for small values of $\alpha$.

Our exploration of the $L_\alpha$s  follows the usual scheme : transfinite induction, and proof-by-cases : zero, successors and limit ordinals.

@@

Let $\beta$ be some ordinal, and assume the arithmetic function $f$ satisfies 
the specification $(\texttt{L\_spec}\;\beta)$.  Let $k$ be any natural number.
Any path from $\texttt{succ}\,\beta$ to $0$ starting at $k+1$ can be decomposed into a first step from $\texttt{succ}\,\beta$ to $\beta$, then a path from
$\beta$ at $k+2$ to $0$. 
By hypothesis the interval $[k+2, f(k+2)-1]$ is $\beta$-mlarge.
But the interval $[k+1, f(k+2)-1]$ is the concatenation of the singleton
$\{k+1\}$ and the interval $[k+2, f(k+2)-1]$.
So, the function $\lambda\,k.\,f(k+1)$ satisfies the specification $\texttt{L\_spec}\,\beta$.

Note that our decomposition of intervals works only if the intervals we consider are not empty. In order to ensure this property, we assume that $f\;k$ is always greater than $k$, which we note \texttt{S <= f}, or \texttt{(fun\_le S f)} (defined 
in~\href{../src/html/hydras.Prelude.Iterates.html\#fun_le}{Prelude.Iterates}).

@@

\section{Using \texttt{Equations}}
\label{sect:L-equations}

@@

\begin{todo}
The proof of \texttt{L\_alpha.L\_ok\_lim} seems to be redundant with
\texttt{Large\_Sets.L\_lim\_ok}.  
\end{todo}

@@

\section{The Wainer-Hardy hierarchy}
\label{sect:hardy}
In order to give an idea of the complexity of the functions  $L_\alpha$s, we compare them with a better known family of functions, the so called \emph{Wainer-Hardy hierarchy} of fast growing functions,
presented for instance in~\cite{Promel2013}. 
\index{Maths!Hardy Hierarchy}

@@

\section{The Wainer hierarchy (functions \texorpdfstring{$F_\alpha$}{F\_alpha})}
\label{sect:wainer}

### 2020-08-12 latex/chapter-termination.tex
proof simplification in L_alpha

\index{Types!T2@T2 (ordinals in Veblen normal form)}
\emph{Please look also at the library
\href{../src/html/hydras.Gamma0.T2.html}{Gamma0.T2.html}}.

@@

The type \texttt{T1} is not a correct ordinal notation, since it contains terms that are not in Cantor normal form. In order to define a type whose inhabitants represent just ordinals, we build a type gathering a term of type \texttt{T1} and a proof that this term is in normal form.
This representation will be used in section~\ref{sect:equations} in order to define recursive functions with the plug-in \texttt{coq-equations}, and also to build an instance of the class \texttt{ON} of ordinal notations.

@@

\index{Coq!Techniques!Type classes}

@@

@@

\begin{Coqsrc}
Instance Epsilon0 : ON Lt compare.  
\end{Coqsrc}

>>>>

\index{Types!T2@T2 (ordinals terms in Veblen normal form)}
\emph{Please look also at the library
\href{../src/html/hydras.Gamma0.T2.html}{Gamma0.T2.html}}.

@@

The type \texttt{T1} is not a correct ordinal notation, since it contains terms that are not in Cantor normal form. Moreover,
in Sections~\ref{sect:equations}, \ref{sect:hardy},
and \ref{sect:wainer}, we needed to define total recursive functions on well-formed ordinal terms less  than $\epsilon_0$, using the \texttt{Equations} plug-in~\cite{sozeau:hal-01671777}.
 In order to define a type whose inhabitants represent just ordinals, we build a type gathering a term of type \texttt{T1} and a proof that this term is in normal form.

@@

\index{Coq!Type classes}

@@

\index{Exercises}
 \begin{exercise}
Prove in \coq{} that for any ordinal $\alpha<\epsilon_0$, $\alpha$ is a limit if 
and only if for all $\beta<\alpha$, the interval $[\beta,\alpha)$ is infinite.
 \end{exercise}

@@

\begin{Coqsrc}
Instance Epsilon0 : ON Lt compare.  
(* ... *)
\end{Coqsrc}
